---
title: "Using a Bayesian Framework for Analysing the Effect of Deprivation on Health Outcomes"
author: "Paul Johnson"
format: 
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---
```{r setup}
#| include: false

library(dplyr)
library(ggplot2)
library(rstanarm)

theme_set(theme_minimal(base_family = "sans"))
bayesplot::color_scheme_set("teal")

deprivation_df <- 
  readr::read_rds(here::here("data", "deprivation.rds"))
```


## Frequentist Vs Bayesian Statistics: A Primer



## Visual Exploration of Deprivation and Health Inequalities Data


```{r imd-score-eda}

summary(deprivation_df$imd_score)

deprivation_df %>%
  ggplot(aes(life_expectancy, imd_score)) +
  geom_point()

```

```{r health-index-eda}

summary(deprivation_df$physiological_risk_factors)

summary(deprivation_df$behavioural_risk_factors)

deprivation_df %>%
  ggplot(aes(life_expectancy, physiological_risk_factors)) +
  geom_point()

deprivation_df %>%
  ggplot(aes(life_expectancy, behavioural_risk_factors)) +
  geom_point()

```

```{r correlations}

deprivation_df %>%
  select(
    life_expectancy,
    physiological_risk_factors,
    behavioural_risk_factors,
    imd_score
         ) %>%
  group_by(area_code) %>%
  summarise_all(.funs = "mean") %>%
  correlation::correlation(bayesian = TRUE) %>%
  summary()

```

There are strong correlations between life expectancy and the three independent variables, however, the correlation between IMD score and the two risk factor variables is also relatively strong, and in the case of behavioural risk factors, it is very strong. This could be an issue.

```{r correlation-eda}

deprivation_df %>%
  ggplot(aes(imd_score, physiological_risk_factors)) +
  geom_point()

deprivation_df %>%
  ggplot(aes(imd_score, behavioural_risk_factors)) +
  geom_point()

deprivation_df %>%
  mutate(imd_decile = as.factor(imd_decile)) %>%
  ggplot(aes(physiological_risk_factors, imd_decile)) +
  ggridges::geom_density_ridges()

deprivation_df %>%
  mutate(imd_decile = as.factor(imd_decile)) %>%
  ggplot(aes(behavioural_risk_factors, imd_decile)) +
  ggridges::geom_density_ridges()

```

The correlations are a little more obvious when plotted and inspected visually. The behavioural risk factors have a positive linear relationship with IMD (as the value of the risk factor goes up, so does IMD score/decile).

## Linear Regression the Frequentist Way

First we will transform each of the explanatory variables to make them a little easier to interpret (particularly with regard to the intercept).

These transformations won't impact the regression results, but will just make the results easier to explain, as the intercept is no longer based on zero values of each explanatory variable (which are effectively meaningless, especially in the Health Index variables).

```{r transforms}

mean_imd <- mean(deprivation_df$imd_score)

deprivation_df <-
  deprivation_df %>%
  mutate(
    imd_transformed = imd_score - mean_imd,
    physiological_transformed = physiological_risk_factors - 100,
    behavioural_transformed = behavioural_risk_factors - 100
    )


```

The frequentist regression is very easy to compute, using `lm()`.

```{r life-expectancy-ols}

life_expectancy_ols <-
  lm(life_expectancy ~ imd_transformed + physiological_transformed + behavioural_transformed,
    data = deprivation_df
    )

summary(life_expectancy_ols)

```

The results suggest that each of the explanatory variables has a small but significant effect on life expectancy. As deprivation increases (IMD score increases), life expectancy decreases, while as the index score representing physiological and behavioural risk factors (meaning a better performance in that Health Index subdomain) increases, life expectancy increases.

## Bayesian Regression

### Constructing Sensible Priors

First we need to come up with some priors. Given everything we already know about life expectancy at birth and the effect that deprivation has on health outcomes, we should be able to constrain our prior distribution in relatively informative ways. 

First, we would expect life expectancy at birth to be normally distributed around 80.

With IMD scores, we can reasonably expect that we should observe a negative relationship on health outcomes as deprivation increases, meaning that as IMD scores increase (meaning higher deprivation), life expectancy should decrease. The effect is unlikely to be huge, so we will conservatively estimate that each one unit increase in IMD score will be associated with a one unit decrease in life expectancy at birth. The key here is more the standard error, which we will constraint sufficiently to mean that our prior is that deprivation will only have a negative effect on life expectancy.

We have similar expectations of the physiological

```{r life-expectancy-priors}
#| echo: true
#| results: hide

life_expectancy_priors <- 
  stan_glm(
    life_expectancy ~ imd_transformed + physiological_transformed + behavioural_transformed,
    data = deprivation_df,
    prior_intercept = normal(80, 1),
    prior = normal(location = c(-1, 1, 1), scale = c(0.25, 0.25, 0.25)),
    prior_PD = TRUE
    )
```

We can see a summary of our priors here:

```{r prior-summary}
prior_summary(life_expectancy_priors)

bayestestR::describe_prior(life_expectancy_priors)
```

The prior distributions look pretty sensible:

```{r prior-plot}
#| message: false
plot(life_expectancy_priors, "hist")

```

## Bayesian GLM

```{r life-expectancy-glm}
#| echo: true
#| results: hide

life_expectancy_glm <- 
  stan_glm(
    life_expectancy ~ imd_transformed + physiological_transformed + behavioural_transformed,
    data = deprivation_df,
    prior_intercept = normal(80, 1),
    prior = normal(location = c(-1, 1, 1), scale = c(0.25, 0.25, 0.25))
    )

life_expectancy_posterior <- as.matrix(life_expectancy_glm)

```

```{r model-diagnostics}

bayestestR::sensitivity_to_prior(life_expectancy_glm)

bayestestR::diagnostic_posterior(life_expectancy_glm)


```

```{r posterior-checks}

bayesplot::ppc_dens_overlay(
  y = life_expectancy_glm$y,
  yrep = posterior_predict(
    life_expectancy_glm,
    draws = 50
    )
  ) +
  labs(
    title = "Posterior Predictive Checks Against Observed Data",
    subtitle = "Samples Drawn from Posterior Predictive Distribution of Life Expectancy\nby IMD Score and Physiological & Behavioural Risk Factors",
    x = "Life Expectancy"
    ) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12)
  )

posterior_vs_prior(
  life_expectancy_glm,
  prob = 0.9,
  group_by_parameter = TRUE,
  facet_args = list(scales = "free")
  ) +
  scale_colour_viridis_d() +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
    )

bayesplot::ppc_scatter_avg(
  y = life_expectancy_glm$y,
  yrep = posterior_predict(
    life_expectancy_glm,
    draws = 50
    )
  )


```

```{r parameter-values}

bayestestR::hdi(
  life_expectancy_glm,
  ci = c(
    0.5, 0.75, 0.89, 0.95
    )
  ) %>%
  plot()


bayestestR::map_estimate(
  life_expectancy_posterior
  ) %>%
  plot()



```